Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-27 paper/paper.qmd
fix grammar error

Instead of examining the impact of abortion laws over broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database, capturing infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a nuanced examination
of the effects of restrictive abortion laws on public health.

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-results, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research.
[Appendix- @sec-data-details] provides external data detail, [Appendix-
@sec-model-details] provides model detail, and [Appendix- @sec-survey]
provides an detailed introduction of surveys, sampling and observational
data in our paper.

@@

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
detailed overview of the Infant Death Rate across the USA. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach
utilizes the interaction between the legality of abortion in the
mother’s state of residence and the timing of the infant’s death as the
predictor variable. Additionally, we include control variables such as
infant year of death, the mother’s state of residence, race, and age to
account for potential confounding factors.

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
@tidyverse was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` @arrow efficiently managed parquet
files for larger datasets. We used `modelsummary` package @modelsummary
to generate clean and interpretable summaries of our regression
models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
clean tables and reproducible reports, and `usmap` for geographical
visualizations. We use `here` @here package simplifies the management of
file paths, ensuring a reproducible workflow by creating a consistent
point of reference for file locations, regardless of the operating
system or working directory.Data cleaning was facilitated by `janitor`
@janitor, while `lubridate` @lubridate handled date-time variables.
Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
regression modeling was conducted using `rstanarm` @rstanarm. Together,
these packages supported an efficient, reproducible workflow for
analyzing and presenting our findings.

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
utilizing comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing public
health strategies and legislative decisions.

@@

```{r}
#| label: fig-deathrate
#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants.The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
#| echo: false

@@

To evaluate the impact of abortion injunctions on infant death rates, we
utilize the Difference-in-Difference (DID) method. As part of this
approach, we construct an interaction term defined as the product of
`abortion_illegal` and `after_injunction`. The resulting interaction
term equals 1 when abortion is illegal and the time is after the
injunction, and 0 otherwise. This term is included in the analysis to
evaluate the joint impact of these factors on infant death rates.

@@

The Age of Mother (`age_of_mother`) refers to the age group of the
mother at the time of childbirth.In our observational data, age is
categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and
30–34 years in the raw data, in compliance with privacy requirements.

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
#| 
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

The self-identified race of the mother, recorded as a single race
category. In this dataset, the available categories are: Asian, Black or
African American, and White. This classification facilitates the
analysis of maternal and infant health outcomes across these racial
groups.

```{r}
#| label: tbl-motherrace
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

```{r}
#| label: fig-choromap
#| fig-cap: Infant Death Rates Across U.S. States (per 1,000 Live Births), December 2022. Infant Mortality Rate is much higher iin Southeastern. Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
#| echo: false

@@

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates.

@@

For our analysis, we use a Bayesian Difference-in-Differences (DID)
model to quantify the impact of abortion injunctions on the infant death
rate across states in the USA. This method compares changes in infant
death rates over time between a treatment group (states where abortion
injunctions were implemented) and a control group (states where abortion
remains legal), providing a robust framework for estimating causal
effects.

@@

In addition to parallel trends, compositional changes in repeated
cross-sectional data over time can influence results. For instance,
shifts in demographic factors such as maternal age or racial makeup
between pre- and post-intervention periods may confound the observed
changes in infant mortality rates. Including control variables helps
adjust for these confounders, ensuring that the estimated effect
reflects the policy intervention rather than changes in population
composition.

@@

The length of the analysis period further complicates the reliability of
results. While longer periods allow for capturing long-term effects,
they also increase the likelihood of unrelated factors, such as new
policies or healthcare advancements, influencing the outcome. Over time,
some states in the control group may also implement similar abortion
policies, diluting the treatment effect. Conversely, shorter analysis
periods

@@

-   **Time Indicator `after_injunction`**: A binary variable indicating
    the time period: 0 for the period before the injunction implemented.
    1 for the period after the injunction implemented.

@@

-   **`state`**: Mother's residence state when the infant born as a
    random effect in the model.

By isolating the impact of the abortion injunction through this
interaction term `after_injunction`×`abortion_illegal`, our model
identifies changes in infant death rates that can be attributed to the
policy intervention. Additional details and diagnostics supporting the
model’s implementation are provided in [Appendix- @sec-model-details].

@@

In our analysis, we employ two models. The first model includes only the
predictor variables and the random effect for state. The second model
builds on the first by incorporating additional control variables to
account for potential confounding factors.

@@

For one of our models, we include control variables (`age_of_mother`,
`mothers_single_race`, and `year_of_death`) to adjust for confounders
that could influence infant death rates, addressing the assumption of
compositional differences in the DID approach and ensuring unbiased
estimates of the intervention’s effect. In the other model, we exclude
these controls to focus on the overall intervention impact, avoid
overfitting, and prioritize simplicity when variability is adequately
accounted for by the design or random effects. Thus, our decision to fit
two models is to compare the outcomes: one adjusts for confounders to
refine the estimates, while the other focuses on the direct effect of
the intervention.

@@

We choose the Bayesian framework since it enhances the DID approach by
incorporating prior information and quantifying uncertainty in a
probabilistic manner. In our Bayesian models, we use weakly informative
priors from the `rstanarm` package, including a Normal prior ($\mu=0$
and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
priors based on data scale. This enhances convergence and ensures priors
are calibrated to predictor ranges without dominating the data. The
model assumes plausible parameter centering, random effects capture
unobserved heterogeneity, and the DID framework relies on parallel
trends between groups. However, limitations include potential bias from
misspecified priors, computational challenges for hierarchical models,
and sensitivity to data sparsity or imbalances. This model may not be
suitable if group-level effects are negligible, parallel trends are
violated, or data is insufficient, but when applied appropriately, it
provides robust and interpretable causal estimates.

@@

modelsummary::msummary(
  list(first_model,second_model),
  fmt = 3,coef_omit="state",notes = "Note. All models include a random variable for state. The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian. Source: CDC Wonder")
```

@@

Demographic variables also exhibit notable associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
coefficient for Black mothers in Model 2 is 6.66, reflecting higher
infant mortality compared to other racial groups.

@@

The analysis reveals distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) reveals that abortion bans during the post-injunction
period result in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a critical role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

The significant interaction between abortion legality and the
post-injunction period illustrates the compounded impact of restrictive
abortion policies on infant mortality rates. These findings underscore
the need for integrated healthcare systems that can address the
consequences of high-risk pregnancies, improve prenatal care, and ensure
equitable access to medical services. Policymakers must consider the
broader public health implications of abortion bans and implement
strategies that safeguard maternal and infant health. This includes
enhancing healthcare delivery in underserved areas, investing in
community-based interventions, and ensuring that restrictive policies
are accompanied by measures to mitigate their public health impact.

@@

In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model. It represent posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data. The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit. In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision. @fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model for infant death rate fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check","Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]

@@

To address these gaps, in this appendix we propose incorporating surveys
to gather qualitative data, which provide rich, detailed accounts of
individual experiences and contextual factors. Qualitative methods allow
for a deeper understanding of how restrictive abortion policies affect
maternal health, access to healthcare, and infant outcomes, capturing
nuances that quantitative analysis alone cannot.

@@

Restrictive abortion laws may exacerbate maternal stress, anxiety, or
depression—especially in cases of unintended or high-risk
pregnancies—potentially undermining maternal health and prenatal care.
Research indicates that maternal mental health is a critical determinant
of infant health outcomes, as conditions such as chronic stress,
depression, or anxiety during pregnancy can negatively impact prenatal
behaviors, access to healthcare, and overall maternal well-being. These
psychological challenges, compounded by societal pressures and systemic
barriers, can significantly increase the risk of adverse infant
outcomes, including infant mortality. The survey will focus on mothers
unable to access abortion services who subsequently experienced the
death of their infant, documenting their experiences to better
understand these interconnected factors.

@@

The objective of this study is to investigate the psychological, social,
and systemic impacts of restrictive abortion laws on maternal health and
infant outcomes. By focusing on mothers unable to access abortion
services who experienced the death of their infant, the study aims to
document lived experiences, explore the role of maternal stress and
healthcare barriers in adverse outcomes, and gather perspectives from
healthcare providers and support organizations. Understanding how
maternal mental health directly affects prenatal care decisions, access
to healthcare, and overall maternal behavior is essential to addressing
the systemic factors contributing to higher infant mortality rates. The
findings will guide the development of evidence-based strategies to
improve maternal mental health, enhance healthcare access, and shape
policy responses to restrictive abortion laws.

@@

In this analysis, we will use non-probability Respondent-driven sampling
to finding respondents. Respondent-driven sampling (RDS), developed by
@heckathorn1997, is ideal for studying hidden populations lacking a
sampling frame and facing potential stigma or harm. Unlike snowball
sampling, RDS offers dual compensation for participation and
recruitment, with recruitment occurring peer-to-peer via social
networks. This method maintains confidentiality, reduces bias through
statistical adjustments, and is particularly effective for sensitive
topics, such as abortion, by utilizing trust within communities.

@@

RDS utilizes peer networks, allowing participants to recruit others they
trust, fostering higher participation rates in populations that may
distrust external researchers. This approach helps maintain
confidentiality and reduces the risk of social stigma. Additionally, RDS
provides a structured framework for statistical adjustments, mitigating
biases inherent in network-based recruitment, and improving the
representativeness of the sample. These advantages make RDS an effective
and ethical choice for gathering insights into the psychological,
social, and systemic impacts of restrictive abortion laws.

@@

We plan to survey 20 respondents for this study. Our sample will consist
of mothers who fit the defined sample frame—those who were unable to
access abortion services due to restrictive laws, experienced the death
of their infant within the first year of life, and are reachable through
Abortion Rights Advocacy Groups. These mothers must also be willing to
participate in and answer our survey questions, ensuring the reliability
and depth of the collected data.

@@

This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias. Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., “Prefer not to say”) to address underreporting in sensitive topics. Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy. Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each sectionof our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.

@@

We draw on recommendations from @stantcheva2023 to minimize response biases. Common response biases identified in survey design include moderacybias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB). Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB.The detailed definitions of these biases are provided in [Appendix-@sec-definition].

@@

3.What is your marital status? - Single - Married - Divorced - Widowed -
Prefer not to say

>>>>

Instead of examining the impact of abortion laws over a broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database, captures infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a nuanced examination
of the effects of restrictive abortion laws on public health.

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-results, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research.
[Appendix- @sec-data-details] provides external data detail, [Appendix-
@sec-model-details] provides model details, and [Appendix- @sec-survey]
provides a detailed introduction to surveys, sampling, and observational
data in our paper.

@@

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from the CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides a
detailed overview of the Infant Death Rate across the USA. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach
utilizes the interaction between the legality of abortion in the
mother’s state of residence and the timing of the infant’s death as the
predictor variable. Additionally, we include control variables such as
infant year of death, the mother’s state of residence, race, and age to
account for potential confounding factors.

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
@tidyverse was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` @arrow efficiently managed parquet
files for larger datasets. We used `modelsummary` package @modelsummary
to generate clean and interpretable summaries of our regression
models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
clean tables and reproducible reports, and `usmap` for geographical
visualizations. We use `here` @here package simplifies the management of
file paths, ensuring a reproducible workflow by creating a consistent
point of reference for file locations, regardless of the operating
system or working directory. Data cleaning was facilitated by `janitor`
@janitor, while `lubridate` @lubridate handled date-time variables.
Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
regression modeling was conducted using `rstanarm` @rstanarm. Together,
these packages supported an efficient, reproducible workflow for
analyzing and presenting our findings.

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
utilizing comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing the public
health strategies and legislative decisions.

@@

```{r}
#| label: fig-deathrate
#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants. The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
#| echo: false

@@

To evaluate the impact of abortion injunctions on infant death rates, we
utilize the Difference-in-Difference (DID) method. As part of this
approach, we construct an interaction term defined as the product of
`abortion_illegal` and `after_injunction`. The resulting interaction
term equals 1 when abortion is illegal the time is after the
injunction, and 0 otherwise. This term is included in the analysis to
evaluate the joint impact of these factors on infant death rates.

@@

The Age of the Mother (`age_of_mother`) refers to the age group of the
mother at the time of childbirth. In our observational data, age is
categorized into the brackets 15–19 years, 20–24 years, 25–29 years, and
30–34 years in the raw data, in compliance with privacy requirements.

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
#| 
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

The self-identified race of the mother was recorded as a single race
category. In this dataset, the available categories are: Asian, Black or
African American, and White. This classification facilitates the
analysis of maternal and infant health outcomes across these racial
groups.

```{r}
#| label: tbl-motherrace
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

```{r}
#| label: fig-choromap
#| fig-cap: Infant Death Rates Across U.S. States (per 1,000 Live Births), December 2022. The infant Mortality Rate is much higher in Southeastern. Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
#| echo: false

@@

@fig-choromap is a choropleth map that visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates.

@@

For our analysis, we use a Bayesian Difference-in-Differences (DID)
model to quantify the impact of abortion injunctions on infant death
rate across states in the USA. This method compares changes in infant
death rates over time between a treatment group (states where abortion
injunctions were implemented) and a control group (states where abortion
remains legal), providing a robust framework for estimating causal
effects.

@@

In addition to parallel trends, compositional changes in repeated
cross-sectional data over time can influence results. For instance,
shifts in demographic factors such as maternal age or racial makeup
between pre-and post-intervention periods may confound the observed
changes in infant mortality rates. Including control variables helps
adjust for these confounders, ensuring that the estimated effect
reflects the policy intervention rather than changes in the population
composition.

@@

The length of the analysis period further complicates the reliability of
the results. While longer periods allow for capturing long-term effects,
they also increase the likelihood of unrelated factors, such as new
policies or healthcare advancements, influencing the outcome. Over time,
some states in the control group may also implement similar abortion
policies, diluting the treatment effect. Conversely, shorter analysis
periods

@@

-   **Time Indicator `after_injunction`**: A binary variable indicating
    the time period: 0 for the period before the injunction was implemented.
    1 for the period after the injunction was implemented.

@@

-   **`state`**: Mother's residence state when the infant was born as a
    random effect in the model.

By isolating the impact of the abortion injunction through this
interaction term `after_injunction`×`abortion_illegal`, our model
identifies changes in infant death rates that can be attributed to the
policy intervention. Additional details and diagnostics supporting the
model’s implementation is provided in [Appendix- @sec-model-details].

@@

In our analysis, we employ two models. The first model includes only the
predictor variables and the random effect for the state. The second model
builds on the first by incorporating additional control variables to
account for potential confounding factors.

@@

For one of our models, we include control variables (`age_of_mother`,
`mothers_single_race`, and `year_of_death`) to adjust for confounders
that could influence infant death rates, addressing the assumption of
compositional differences in the DID approach and ensuring unbiased
estimates of the intervention’s effect. In the other model, we exclude
these controls focus on the overall intervention impact, and avoid
overfitting, and prioritize simplicity when variability is adequately
accounted for by the design or random effects. Thus, our decision to fit
two models to compare the outcomes: one adjusts for confounders to
refine the estimates, while the other focuses on the direct effect of
the intervention.

@@

We choose the Bayesian framework since it enhances the DID approach by
incorporating prior information and quantifying uncertainty in a
probabilistic manner. In our Bayesian models, we use weakly informative
priors from the `rstanarm` package, including a Normal prior ($\mu=0$
and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
priors based on the data scale. This enhances convergence and ensures priors
are calibrated to predictor ranges without dominating the data. The
model assumes plausible parameter centering, random effects capture
unobserved heterogeneity, and the DID framework relies on parallel
trends between groups. However, limitations include potential bias from
misspecified priors, computational challenges for hierarchical models,
and sensitivity to data sparsity or imbalances. This model may not be
suitable if group-level effects are negligible, parallel trends are
violated, or data is insufficient, but when applied appropriately, it
provides robust and interpretable causal estimates.

@@

modelsummary::msummary(
  list(first_model,second_model),
  fmt = 3,coef_omit="state",notes = "Note. All models include a random variable for the state. The reference level of age_of_mother is 15-19. The reference level of monthers_single_race is Asian. Source: CDC Wonder")
```

@@

Demographic variables also exhibit notable associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
the coefficient for Black mothers in Model 2 is 6.66, reflecting a higher
infant mortality compared to other racial groups.

@@

The analysis reveals distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) reveals that abortion bans during the post-injunction
period results in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a critical role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

The significant interaction between abortion legality and the
post-injunction period illustrates the compounded impact of restrictive
abortion policies on infant mortality rates. These findings underscore
the need for integrated healthcare systems that can address the
consequences of high-risk pregnancies, improve prenatal care and ensure
equitable access to medical services. Policymakers must consider the
broader public health implications of abortion bans and implement
strategies that safeguard maternal and infant health. This includes
enhancing healthcare delivery in underserved areas, investing in
community-based interventions, and ensuring that restrictive policies
are accompanied by measures to mitigate their public health impact.

@@

In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model. It represents posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data. The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit. In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision. @fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model for infant death rate fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check", "Second Model Posterior prediction check", "First Model Comparing the posterior with the prior", "Second Model Comparing the posterior with the prior"]

@@

To address these gaps, in this appendix we propose incorporating surveys
to gather qualitative data, which provide rich, detailed accounts of
individual experiences and contextual factors. Qualitative methods allow
for a deeper understanding of how restrictive abortion policies affect
maternal health, access to healthcare, and infant outcomes capturing
nuances that quantitative analysis alone cannot.

@@

Restrictive abortion laws may exacerbate maternal stress, anxiety, or
depression—especially in cases of unintended or high-risk
pregnancies—potentially undermining maternal health and prenatal care.
Research indicates that maternal mental health is a critical determinant
of infant health outcomes, as conditions such as chronic stress,
depression, or anxiety during pregnancy can negatively impact prenatal
behaviors, access to healthcare, and overall maternal well-being. These
psychological challenges, compounded by societal pressures and systemic
barriers can significantly increase the risk of adverse infant
outcomes, including infant mortality. The survey will focus on mothers
unable to access abortion services who subsequently experienced the
death of their infant, documenting their experiences to better
understand these interconnected factors.

@@

The objective of this study is to investigate the psychological, social,
and systemic impacts of restrictive abortion laws on maternal health and
infant outcomes. By focusing on mothers unable to access abortion
services who experienced the death of their infant, the study aims to
document lived experiences, explore the role of maternal stress and
healthcare barriers in adverse outcomes, and gather perspectives from
healthcare providers and support organizations. Understanding how
maternal mental health directly affects prenatal care decisions, access
to healthcare, and overall maternal behavior is essential to addressing
the systemic factors contributing to higher infant mortality rates. The
findings will guide the development of evidence-based strategies to
improve maternal mental health, enhance healthcare access and shape
policy responses to restrictive abortion laws.

@@

In this analysis, we will use non-probability Respondent-driven sampling
to find respondents. Respondent-driven sampling (RDS), developed by
@heckathorn1997, is ideal for studying hidden populations lacking a
sampling frame and facing potential stigma or harm. Unlike snowball
sampling, RDS offers dual compensation for participation and
recruitment, with recruitment occurring peer-to-peer via social
networks. This method maintains confidentiality, reduces bias through
statistical adjustments, and is particularly effective for sensitive
topics, such as abortion, by utilizing trust within communities.

@@

RDS utilizes peer networks, allowing participants to recruit others they
trust, and fostering higher participation rates in populations that may
distrust external researchers. This approach helps maintain
confidentiality and reduces the risk of social stigma. Additionally, RDS
provides a structured framework for statistical adjustments, mitigating
biases inherent in network-based recruitment, and improving the
representativeness of the sample. These advantages make RDS an effective
and ethical choice for gathering insights into the psychological,
social, and systemic impacts of restrictive abortion laws.

@@

We plan to survey 20 respondents for this study. Our sample will consist
of mothers who fit the defined sample frame—those who were unable to
access abortion services due to restrictive laws, experience death
of their infant within the first year of life, and are reachable through
Abortion Rights Advocacy Groups. These mothers must also be willing to
participate in and answer our survey questions, ensuring the reliability
and depth of the collected data.

@@

This survey explores the impacts of restrictive abortion laws on maternal mental health and infant outcomes, emphasizing neutral, carefully crafted questions to ensure participant comfort and minimize social desirability bias. Inspired by @lindberg2022, it employs strategies like thoughtful question placement, phrasing, and opt-out options (e.g., “Prefer not to say”) to address underreporting in sensitive topics. Combining multiple-choice and open-ended formats, the survey aims to balance structured data collection with nuanced insights, fostering trust and enhancing data accuracy. Also, based on @lindberg2022 findings that respondents preferred introductions emphasizing how sharing their abortion experiences could help improve health services for other women, we adopted a “helping” framework in each section of our survey introductions to encourage accurate reporting and increase participants’ motivation to contribute.

@@

We draw on recommendations from @stantcheva2023 to minimize response biases. Common response biases identified in survey design include moderacy bias, extreme response bias, ordering bias, acquiescence bias, experimenter demand effect(EDE), and social desirability bias (SDB). Our survey primarily focuses on strategies to reduce moderacy bias, extreme response bias, ordering bias, and SDB. The detailed definitions of these biases are provided in [Appendix-@sec-definition].

@@

3. What is your marital status? - Single - Married - Divorced - Widowed -
Prefer not to say
